Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India

### 2025-08-04 manuscript.Rmd
Edit manuscript according to AJE proofreading

**Objective** To assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes.

**Design** A pilot pragmatic three-armed parallel, cluster randomised, controlled trial between `r start.date` and `r end.date`. Patient follow up was 30 days.

**Setting** Tertiary care hospitals across metropolitan areas in India.

**Participants** Adult trauma patients and residents managing these patients.

**Interventions** ATLS or PTC training for residents in the intervention arms.

**Main Outcomes and Measures** The outcomes were consent rate, lost to follow up rate, missing data rates, differences in distribution between observed and data extracted from medical records, and resident pass rate.

**Results** Two hospitals were randomised to ATLS, two to PTC, and three to standard care. We included `r n.patients` patients and `r n.residents` residents. The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. The pass rate was `r pass.rate`%. The missing data was overall low for key variables. Data collected through observations were similar to data extracted from medical records, but there was more missing data in the extracted data.

**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes will be feasible, especially if such a trial would use data and outcomes available in medical records.

@@

- Prospective data collection with direct observations by dedicated research officers.
- Lack of a priori defined success criteria and thresholds for the feasibility outcomes.
- Use of sealed envelopes potentially compromised allocation concealment.
- Participating centers' heterogeneity may affect the study estimates and bias the results.

@@

Trauma, defined as the clinical entity composed of physical injury and the body's associated response, causes 4.3 millions deaths every year [@injuries2020]. Several trauma life support training programs have been developed to improve the early management of patients in hospital by providing a structured framework to assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].

The proprietary Advanced Trauma Life Support (ATLS) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].

Several studies, including at least two randomised studies [@Ali1996; @Ali1999], show that ATLS is associated with improved knowledge and skills among providers [@Mohammad2013]. Observational evidence suggests that PTC also leads to improved provider skills [@Kadhum2020]. The missing link is then whether these improved knowledge and skills translate into measurably improved patient outcomes.

Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible. Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.

We therefore conducted a pilot study to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS and PTC with standard care on outcomes in adult trauma patients.

@@

We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. There were a standard care arm and two intervention arms, ATLS and PTC training. We planned to collect data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). The actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters. We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline. In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [@teresi_guidelines_2022].

@@

We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS, PTC, nor any other established trauma life support training program is routinely taught. Details about each cluster is included as Supplementary Table S1. Originally, we intented to include six hospitals as clusters, but decided to include a seventh when they expressed interest and as we had the budget to accommodate this. These seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.

@@

We defined a cluster as a tertiary care hospitals in metropolitan areas in India that admitted more than 400 adult patients with trauma each year, and that had operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock. In each cluster, we trained one or more units of physicians providing trauma care in the emergency department. To be eligible, units could have no more than 25% of their physicians trained in either ATLS, PTC, or similar training programmes before the start of the pilot study. Those residents who had received training in the last five years were considered as trained. The figure of 25% was decided through consensus in the research team, to balance feasibility and contamination of results. The principal investigator at each hospital selected the units for training. We randomised on the hospital level to avoid contamination between intervention arms and the standard care arms.

@@

We trained resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training. Consent was sought from the residents in each of the intervention groups before they underwent the ATLS or PTC training. In the published protocol, we stated that only surgical residents would be trained. However, in some of the participating hospitals, emergency medicine residents performed the initial resuscitation and management of trauma patients, and we therefore decided to train them instead.

@@

We included persons who were 15 years or older and presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty. History of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter 20 of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting.

@@

Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units. These units include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department. In the emergency department, trauma patients are initially assessed by residents in these units, and they also resuscitate patients, perform interventions and refer patients for imaging or other investigations. Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management. We did not collect data on how standard care varied between the participating hospitals.

@@

In each intervention arm the residents in one or two units were trained in either ATLS or PTC at the beginning of the three month intervention phase. For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit. We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients. The ATLS training was conducted in an ATLS certified training centre in Mumbai and the PTC training was conducted in New Delhi. Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.

The provider courses of both programmes are two days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations. The teaching is based on case discussions and skill stations. There are some important differences between the two programmes. The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018]. The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018]. The ATLS programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK based charity the PTC Foundation and is free of charge.

## Feasibility outcomes

Our feasibility outcomes were:

- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all included patients.
- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
- Differences in distributions between directly observed data and data extracted from medical records. Distribution refers to summary statistics and directly observed data refers to data collected by project officers while observing the delivery of care. This outcome applied to all variables that could be reasonably expected to be present in the medical records, but we decided to only extract this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.

We did not prespecify criteria to judge whether to continue with the full scale trial.

@@

## Sample size

We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training. We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods. We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.

@@

Arriving patients were screened for eligibility and consented, if conscious. Unconscious patients were consented by the patient’s representative. This proxy consent was reaffirmed by the patient, on regaining consciousness. Patients who did not regain consciousness were included based on their representative's consent. We followed up patients at 24 hours after arrival at the emergency department, and at 30 days after arrival at the emergency department. The follow up period for each patient was therefore one month.

@@

Participating units were screened for eligibility once hospitals confirmed their participation. All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms. The protocol stated that residents would be approached for consent before the hospitals were randomised, but this proved not to be feasible, and we therefore asked residents for consent after the hospitals were randomised but before training. The training was conducted approximately one month after the study started in that hospital. Initially, we planned to use simple random sampling to select the units to be trained in each intervention hospital, but for pragmatic reasons we decided to leave the decision on which units to train to the site principal investigator. The number of residents trained in each intervention cluster varied because the size of the units varied.

## Allocation and blinding

We used simple randomisation implemented using sealed envelopes to allocate sites to trial arms. It was not possible to blind investigators, residents or patients to the intervention. We did not blind the data analysts.

@@

The planned data collection period was four months, but the actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters. A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care. The research officers observed care and interviewed residents and patients, and also extracted data from the hospital records. We followed up admitted patients for their complications and other in-hospital outcome measures. Patients who were not admitted or who were discharged before the end of the study were followed up telephonically for mortality outcomes and quality of life outcomes.

@@

The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these ICD-10 codes, we calculated the Injury Severity Score (ISS) using the R package `icdpicr` [@icdpicr_2022]. The ISS is a widely used measure of injury severity and ranges from 0 to 75, with a cutoff of 16 often being used to define major trauma, and 75 representing unsurvivable trauma. We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and health related quality of life (using the EQ-5D-3L). We did not calculate an EQ-5D-3L index score, because there is currently no Indian value set available for the EQ-5D-3L [@jyani_development_2022]. We also attempted to collect data on cause of death. A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

## Patient and public involvement

We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial. The results of these consultations are published separately [@david_patient-reported_2024]. We initially planned to distribute periodic surveys to residents and follow them up after 30 days after training, but we changed this to interviews after the end of the study period as we thought this would provide richer data (not published).

@@

## Data monitoring

We conducted weekly online meetings to monitor the study and data collection. We conducted one interim analysis approximately halfway through the study, and decided to complete the study as residents and patients were consenting to be included in the study and key variables including mortality outcomes could be collected. We did not use a data monitoring committee.

@@

We used the `r version$version.string` Statistical Software for all analyses [@R_2024]. We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014]. Initially, we planned to analyse feasibility outcomes both on an overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results. Quantitative variables are summarised as median and interquartile range. Qualitative variables are presented as absolute numbers and percentages. Additional analyses performed according to the original protocol are available as additional online material [@warnberg_additional_2025].

@@

We were granted research ethics approval from the institutional ethics committees at each participating hospital. For each participating hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, MC/KOL/IEC/NON-SPON/1217/11/21 for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for the Postgraduate Institute of Medical Education and Research, Chandigarh.

@@

We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`. The period of data collection and the number of patients included per month per cluster is shown in Figure \@ref(fig:patients-per-month-figure). Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively. The standard care arm included `r n.control` patients, the ATLS arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients. We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS, and  `r number_to_text(n.ptc.residents)` in PTC.

```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month. Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively."}
knitr::include_graphics("patients-per-month.png")
```

The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics). Extended patient sample characteristics are shown in Supplementary Table S2. Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`. These prognostic factors differed between the trial arms. A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.

After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS arm and `r m30d.ptc` patients in the PTC arm. The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively.

@@

The percentage of patients who consented to follow up was `r p.consent`% and the lost to follow up rate was `r rate.lost.to.follow.up`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2. The variables with the maximum amount of missing data were  the cost of treatment, complications and cause of death, also reported in Supplementary Tables S2.

The differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview, are shown in Table \@ref(tab:observed-vs-extracted). Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records compared to the directly observed data.

The percentage of residents who consented to training was `r recruitment.rate.residents`% and the pass rate was `r pass.rate`%.

```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview in a convenience sample of patients."}
observed.vs.extracted.comparison.table
```

@@

We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete. To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.

@@

All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients. This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge. Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials. Following up patients after discharge is notoriously challenging in the setting of this study, and the full scale trial might need to focus on in-hospital mortality as the primary outcome.

@@

During the course of this pilot study we deviated from the protocol in several ways. The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes. However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [@eldridge_how_2015; @teresi_guidelines_2022]. We therefore decided to only report patient outcomes descriptively. Another important deviation from the protocol was the training of emergency medicine residents. We initially planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we decided to adapt to local routines. The full scale study will need to accommodate this variation as part of the protocol.

>>>>

**Objective** The aim was to assess the feasibility of conducting a cluster randomised controlled trial comparing the effects of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes.

**Design** This was a pilot pragmatic three-armed parallel, cluster randomised, controlled trial conducted between `r start.date` and `r end.date`. Patient were followed up for 30 days.

**Setting** The study setting was tertiary care hospitals across metropolitan areas in India.

**Participants** Adult trauma patients and residents managing these patients were included.

**Interventions** ATLS or PTC training was provided for residents in the intervention arms.

**Main Outcomes and Measures** The outcomes were the consent rate, lessons to follow up rate, missing data rates, differences in the distribution between observed data and data extracted from medical records, and the resident pass rate.

**Results** Two hospitals were randomised to the ATLS arm, two to the PTC arm, and three to the standard care arm. We included `r n.patients` patients and `r n.residents` residents. The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to receive training was `r recruitment.rate.residents`%. The loss to follow up rate was `r rate.lost.to.follow.up`%. The pass rate was `r pass.rate`%. Overall, the amount of missing data for key variables was low. The data collected through observations were similar to data extracted from medical records, but there were more missing extracted data.

**Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes appears feasible, especially if such a trial would use data and outcomes available in medical records.

@@

- Prospective data collection with direct observations made by dedicated research officers.
- A lack of a priori defined success criteria and thresholds for feasibility outcomes.
- The use of sealed envelopes potentially compromised allocation concealment.
- Heterogeneity of the participating centers may affect the study estimates and introduce bias.

@@

Trauma, defined as a clinical entity composed of physical injury and the associated response of the body, causes 4.3 million deaths every year [@injuries2020]. Several trauma life support training programmes have been developed to improve the early management of patients in hospitals by providing a structured framework for assessment and treatment [@Mohammad2013; @Jayaraman2014; @Kadhum2020].

The proprietary Advanced Trauma Life Support (ATLS) and low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].

Several studies, including at least two randomised studies [@Ali1996; @Ali1999], have shown that ATLS is associated with improved knowledge and skills among providers [@Mohammad2013]. Observational evidence also suggests that PTC also leads to improved provider skills [@Kadhum2020]. The missing link is whether this improved knowledge and skill set translates into measurably improved patient outcomes.

Systematic reviews have called for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], as conducting such effectiveness trials in settings where they are part of the standard of care is not feasible. Many of these settings are in low- and middle income countries, where trial logistics can be more challenging.

We therefore conducted a pilot study with the aim to assess the feasibility of a cluster randomised controlled trial comparing the effects of ATLS and PTC with standard care on outcomes in adult trauma patients.

@@

We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. The trial included a standard care arm and two intervention arms, ATLS and PTC training. We planned to collect data for four months in all three arms, first during a one month observation phase, followed by a three-month intervention phase (or continued observation in the standard care arm). The actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection in the intervention clusters post-training. We included a one month observation phase to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in changes from baseline. In the published protocol, we also aimed to estimate probable effect sizes and other parameters needed for sample size calculations for a full-scale trial [@GerdinWärnberg2022]. However, we revised this aim in light of current guidance on the conduct and reporting of pilot trials [@teresi_guidelines_2022].

@@

We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS, PTC, nor any other established trauma life support training program is routinely taught. Details about each cluster are provided in Supplementary Table S1. We initially intented to include six hospitals as clusters, but added a seventh hospital when management expressed interest, as we had the budget to accommodate the request. These seven hospitals represented a convenience sample that fulfilled the inclusion criteria, and had existing connections to the research team.

@@

We defined a cluster as a tertiary care hospital in metropolitan areas in India that admit more than 400 adult patients with trauma annually, and has operation theatres, X-ray, CT, and ultrasound facilities, and blood bank available around the clock. In each cluster, we trained one or more units of physicians providing trauma care in the emergency department. To be eligible, units could have no more than 25% of their physicians with previous training in either ATLS, PTC, or similar training programmes. Residents who had received training in the last five years were considered trained. The 25% threshold was determined through consensus within the research team, to balance feasibility and the risk of contamination. The principal investigator at each hospital selected the units for training. We randomised at the hospital level to avoid contamination between the intervention and standard care arms.

@@

We trained resident doctors undertaking speciality training in surgery or emergency medicine who managed trauma patients in the emergency department, and were expected to remain in the participating hospitals for at least one year after training. Consent was obtained from the residents in each intervention arm before ATLS or PTC training. In the published protocol, we stated that only surgical residents would be trained. However, in some of the participating hospitals, emergency medicine residents led the initial resuscitation and management of trauma patients, and we therefore included them in the training.

@@

We included persons aged 15 years or older who presented to the emergency department at participating hospitals with a history of trauma when a designated unit was on duty. A history of trauma was defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter 20 of the International Classification of Disease version 10 (ICD-10) codebook as the reason for presentation.

@@

Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians into units. These units include both faculty members and residents, who are assigned a specific day of the week to work in the emergency department. Trauma patients are initially assessed by residents in these units, who also resuscitate patients, perform interventions and refer patients for imaging or other investigations. Compared with settings that adopt a trauma team approach, nurses and other healthcare professionals are involved to a limited extent during initial management. We did not collect data on how standard care varied between the participating hospitals.

@@

In each intervention arm, residents from one or two units were trained in either ATLS or PTC at the beginning of the three month intervention phase. For the purpose of this pilot study, our target was to train at least 75% of residents in each unit. Faculty members were not trained, because they are typically not directly involved in the initial management of trauma patients. ATLS training was conducted at an ATLS certified training centre in Mumbai and PTC training was conducted in New Delhi. Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.

The provider courses of both programmes take place across two days, and focus on the assessment, resuscitation and stabilisation of trauma patients, with adaptions for different patient populations. Teaching is based on case discussions and skill stations. There are several important differences between the two programmes. The ATLS course focuses more on inter-hospital patient transfers and includes a greater emphasis on the trauma team [@acsAtls2018]. In contrast, the PTC course focuses on trauma care in the low resource setting [@ptcfPtc2018]. The ATLS programme is run by the American College of Surgeons and requires a participant fee, whereas the PTC programme is run by the UK-based PTC Foundation and is provided free of charge.

## Feasibility Outcomes

Our feasibility outcomes were as follows:

- Consent rates of patients and residents. This was defined as the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Loss to follow-up rate. This applied only to patients and was defined as the percentage of patients among all included patients who did not complete the 30 day follow up.
- Missing data rate. This applied to each outcome and variable and was defined as the percentage of missing values.
- Differences in distributions between directly observed data and data extracted from medical records. Distribution refers to summary statistics and directly observed data refers to data collected by project officers while observing the delivery of care. This outcome applied to all variables that could be reasonably expected to be present in the medical records. To reduce workload, these data were extracted from a convenience sample of patients only.
- Pass rate. This applied only to residents in the intervention arms and was defined as the percentage of residents who passed the training programme, among all residents who received training.

We did not prespecify criteria to determine whether to proceed with a full-scale trial.

@@

## Sample Size

We aimed to include at least two clusters per arm to avoid basing conclusions on single centres. We also aimed to train at least two units per intervention cluster to evaluate the logistics of sending residents for training. We did not conduct a formal power calculation for this pilot, as the primary purpose of the study was to assess the feasibility of the trial logistics and research methods. We anticipated variation in the number of patients included per cluster depending on hospital patient volume.

@@

Incoming patients were screened for eligibility and consented, if they were conscious and able to provide consent. For unconscious patients, consent was provided by a patient representative. These patients reaffirmed this proxy consent, upon regaining consciousness. Patients who did not regain consciousness remained included based on their representative's consent. We followed up patients at 24 hours and at 30 days after arrival at the emergency department. The follow-up period for each patient was therefore one month.

@@

Participating units were screened for eligibility once the hospitals confirmed participation. All residents in these units were approached for consent to training if their hospital was randomised to one of the intervention arms. The protocol stated that residents would be approached for consent before randomisation, but this proved not to be feasible. Instead, we asked residents for consent after the hospitals were randomised but before training. Training took place approximately one month after study initiation in that hospital. We initially planned to use simple random sampling to select the units to be trained, but for pragmatic reasons this decision was left to the site principal investigator. The number of residents trained in each intervention cluster varied based on the unit size.

## Allocation and Blinding

We used simple randomisation implemented using sealed envelopes to allocate sites to the trial arms. It was not possible to blind investigators, residents or patients to the intervention. Data analysts were not blinded.

@@

The planned data collection period was four months. However, the actual period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters. Research officers collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care. The research officers observed care, interviewed residents and patients, and extracted data from hospital records. Admitted patients were followed up for complications and other in-hospital outcome measures. Patients who were not admitted or who were discharged before the end of the study were followed up by telephone for mortality and quality-of-life.

@@

Research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these codes, we calculated the Injury Severity Score (ISS) using the R package `icdpicr` [@icdpicr_2022]. The ISS is a widely used measure of injury severity and ranges from 0 to 75, with a cut-off score of 16 often used to define major trauma and 75 representing unsurvivable trauma. We also collected data on potential outcomes for the full-scale trial, including 30-day and in-hospital mortality, complications and health related quality of life (assessed using the EQ-5D-3L). We did not calculate an EQ-5D-3L index score, because no Indian value set is currently available [@jyani_development_2022]. We also attempted to collect data on cause of death. A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

## Patient and Public Involvement

We conducted community consultations with patients, their caregivers, patient groups, and resident doctors to inform the selection of outcome measures and implementation of the full-scale trial. Results of these consultations are published separately [@david_patient-reported_2024]. We initially planned to distribute periodic surveys to residents and follow them up 30 days after training, but this was later changed to end-of-study interviews to allow for richer data (not published).

@@

## Data Monitoring

Weekly online meetings were held to monitor study progress and data collection. One interim analysis was conducted approximately halfway through the study, and it was decided to complete the study, as residents and patients were consenting to be included in the study and key variables including mortality, could be collected. A formal data monitoring committee was not used.

@@

All analyses were conducted using `r version$version.string` statistical software [@R_2024]. Feasibility outcomes and other data were analysed using descriptive statistics and no formal hypothesis testing was performed. Initially, we planned to analyse feasibility outcomes at both the overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results. Quantitative variables are summarised as medians and interquartile ranges. Qualitative variables are presented as absolute numbers and percentages. Additional analyses performed according to the original protocol are available as additional online material [@warnberg_additional_2025].

@@

We were granted research ethics approval from the institutional ethics committees at each participating hospital. For each hospital, the approvals were HBTMC/266/SURGERY for Dr R N  Cooper Municipal General Hospital in Mumbai, IEC(II)/OUT/134/2022 for Seth GS Medical College and KEM Hospital in Mumbai, ICC/214/22/20/05/2022 for Lokmanya Tilak Municipal Medical College and General Hospital, CREC/2022/FEB/1(ii) for MEDICA Superspeciality Hospital in Kolkata, MC/KOL/IEC/NON-SPON/1217/11/21 for Medical College, Kolkata, NRSMC/IEC/93/2021 for Nilratan Sircar Medical College & Hospital in Kolkata, and finally IEC-03/2022-2332 for the Postgraduate Institute of Medical Education and Research, Chandigarh.

@@

We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`. The  data collection period and the number of patients included per month per cluster are shown in Figure \@ref(fig:patients-per-month-figure). Owing to an error in the data uploading process, data were available only for one and three months in two clusters respectively. The standard care arm included `r n.control` patients, the ATLS arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients. A total of  `r number_to_text(n.residents)` residents were trained, `r number_to_text(n.atls.residents)` in ATLS and  `r number_to_text(n.ptc.residents)` in PTC.

```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month. Owing to an error in the data uploading process, data were available for only one and three months in two clusters respectively."}
knitr::include_graphics("patients-per-month.png")
```

The study flow diagram is shown in Figure \@ref(fig:flow-diagram), and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics). Extended patient sample characteristics are shown in Supplementary Table S2. Overall, `r females` patients were females, the median (interquartile range, IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`. These prognostic factors differed between the trial arms. A total of `r m30d` patients died within 30 days of arrival at the emergency department, and `r inhosp` patients died in the hospital.

After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared with `r m30d.atls` patient in the ATLS arm and `r m30d.ptc` patients in the PTC arm. The corresponding rates for in-hospital mortality were `r inhosp.standard.care`, `r inhosp.atls`, and `r inhosp.ptc` for the standard care, ATLS and PTC arms, respectively.

@@

The percentage of patients who consented to follow-up was `r p.consent`% and the percentage lost to follow-up was `r rate.lost.to.follow.up`%. The missing data rate ranged from `r min.missing.data`% to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2. The variables with the largest amount of missing data were  the cost of treatment, complications and cause of death, also reported in Supplementary Table S2.

Differences in distributions between directly observed data and data extracted from medical records for selected variables collected through observation or interviews are shown in Table \@ref(tab:observed-vs-extracted). Overall, the data were similarly distributed, but there were considerably more missing values in the data extracted from medical records than in the directly observed data.

The percentage of residents who consented to training was `r recruitment.rate.residents`% and the pass rate was also `r pass.rate`%.

```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables collected through observation or interview in a convenience sample of patients."}
observed.vs.extracted.comparison.table
```

@@

We demonstrated that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS with PTC and standard care. The missing data rate was low for key variables. However, some variables had very high missing data rates and may not be feasible to include in a full-scale trial, or may require different data collection methods. The missing data rate was substantially higher when data were extracted from medical records rather than directly observed, although the distributions were similar. Thus suggests that data collected from medical records are reliable even if they are less complete. To increase the completeness of data extracted from the medical records, a full-scale trial should limit the number of variables extracted from medical records and emphasise the importance of having these variables recorded to the participating hospitals.

@@

All-cause 30-day mortality data were missing for `r rate.lost.to.follow.up`% of patients. This rate may be high, especially compared to, for example, the CRASH-2 and REACT-2 trials, which reported missing primary outcomes for fewer than 0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow patients after discharge. Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to those in previous trials. Following patients after discharge is notoriously challenging in this setting, and the full-scale trial may need to focus on in-hospital mortality as the primary outcome.

@@

During this pilot study we deviated from the protocol in several ways. The most significant deviation was a revision of the study aim, as we initially intended to estimate potential effect sizes and other parameters to help sample size calculations for a full-scale trial, in addition to assessing the feasibility outcomes. However, current guidance advises against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [@eldridge_how_2015; @teresi_guidelines_2022]. We therefore chose to report patient outcomes descriptively. Another significant deviation was the training of emergency medicine residents. We originally planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we adapted to local routines. A full-scale study will need to accommodate this variation as part of the protocol.
